### 2020-03-04 manuscript.tex
Clock model averaging

In this article we consider application of the AVMVN kernel to the branch rates of leaf nodes. This operator is not readily applicable to internal node branch rates due to their dependency on the topology of the tree.

@@

\begin{align}
	\mathcal{AVMVN}(x) = (1-\beta) \mathcal{MVN}\big(x, \frac{\Sigma_N}{N} \big) + \beta \mathcal{MVN}\big(x, \frac{\mathbb{I}_N}{N} \big),
\end{align}

where $\mathcal{MVN}$ is the multivariate normal probability density. $\beta \; (= 0.05)$ is a constant which determines the fraction of the proposal determined by the identity matrix ${I}_N$, as opposed to the covariance matrix $\Sigma_D$ which is trained during MCMC.

@@

@@

where $\hat{L}$ is the maximim likelihood and $P(\hat{\sigma})$ is the prior density of $\sigma$. This analysis reveals that when there are a small number of rates, data simulated from a lognormal distribution is often best modelled with an exponential distribution, especially when $\sigma$ is close to 1 (\text{Fig. \ref{fig:ClockModelAverging}}). However, for larger trees $N > 30$, clock model averaging should be a viable approach.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/ClockModelAverging.pdf}
\caption{\textbf{Ability to resolve LogNormal and Exponential distributed data.} \textit{Left:} at each grid point, $2N-2$ random variables are sampled from a $\text{LogNormal}(-0.5\sigma^2, \sigma)$ distribution. The point is assessed for its fit to i) a $\text{LogNormal}(-0.5\hat{\sigma}^2, \hat{\sigma})$ distributon, and ii) an $\text{Exponential}(1)$ disribution using the Bayesian information criterion (BIC). This is repeated 1000 $\times$ at each grid point and the two sets of BIC scores are compared using a paired t-test ($p=0.001$), which determines the colour of the point on the plot. \textit{Right:} points are instead sampled from an $\text{Exponential}(1)$ disribution. These simulations show that clock model averaging is most applicable for trees with more than 30 taxa.   } 
\label{fig:ClockModelAverging}
\end{figure}

@@

To avoid a cross-product explosion, the three targets for clock model improvement (\textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, and \textbf{\nameref{sect:NER}}) are evaluated sequentially, in the order presented in this paper. The setting(s) which are considered to be the best in each step are then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig. \ref{fig:tournament}}.

@@

\textbf{Fig. \ref{fig:parameterisationResults}} shows that the \textit{real 2006} performs considerably worse than any of the other settings. This is due to the poor sampling of the prior under this setting (ie. low ESS of $p(\theta)$). The failure of \textit{real 2006} thus highlights the appeal of 1) the \textit{cat} or \textit{quant} parameterisations, both of which have trivial contributions to the prior density (ie. uniform priors), and 2) the smarter operators used by \textit{real} (Zhang and Drummond 2020). Due to its computational burden, \textit{real 2006} was not benchmarked for all of the datasets in \textbf{Table \ref{table:datasets}}.

%Our results are also consistent with Zhang and Drummond 2020, who found that \textit{real} (with constant distance operators) gives superior mixing of rates compared with \textit{cat}.

Our results show that the \textit{quant} parameterisation yields the best performance with respect to effective samples per hour. \textit{quant} outperforms \textit{quant 2006}, suggesting that the constant distance operators are effective. Furthermore, \textit{quant} outperforms \textit{real} especially at sampling from the posterior and prior distributions (ie. high ESS/hr for $p(\theta|D)$ and $p(\theta)$). This is most likely because of the uniform prior distribution of rate quantiles.

Overall, \textit{quant} yields a median ESS/hr approximately 150 \% faster with respect to sampling the posterior probability, and approximately 30 \% faster with respect to sampling branch rates $r$ and clock standard deviation $\sigma$, compared with \textit{real}.

%\text{cat} and \textit{quant 2006} parameterisations yield similar effective samples per hour for all parameters (including posterior, likelihood, and prior probabilities).

@@

\subsubsection*{Clock model averaging}

>>>>

In this article we consider application of the AVMVN kernel to the branch rates of leaf nodes. This operator is not readily applicable to internal node branch rates due to their dependencies on tree topology.

@@

\begin{align}
	\mathcal{AVMVN}(x) =  \mathcal{MVN}\big(x, (1-\beta) \frac{\Sigma_N}{N} + \beta \frac{\mathbb{I}_N}{N} \big) ,
\end{align}

where $\mathcal{MVN}$ is the multivariate normal probability density. $\beta \; (= 0.05)$ is a constant which determines the fraction of the proposal determined by the identity matrix $\mathbb{I}_N$, as opposed to the covariance matrix $\Sigma_D$ which is trained during MCMC.

@@

\subsection*{Clock model averaging}
\label{sect:clockModelAveraging}

@@

where $\hat{L}$ is the maximim likelihood and $P(\hat{\sigma})$ is the prior density of maximum likelihood estimate $\hat{\sigma}$. This analysis reveals that when there are a small number of rates, data simulated from a lognormal distribution is often best modelled with an exponential distribution, especially when $\sigma$ is close to 1 (\text{Fig. \ref{fig:ClockModelAverging}}). However, for larger trees $N > 30$, clock model averaging should be a viable approach.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/ClockModelAverging.pdf}
\caption{\textbf{Ability to resolve LogNormal and Exponential distributed data.} \textit{Left:} at each grid point, $2N-2$ random variables are sampled from a $\text{LogNormal}(-0.5\sigma^2, \sigma)$ distribution. The samples are assessed for their fit to i) a $\text{LogNormal}(-0.5\hat{\sigma}^2, \hat{\sigma})$ distributon, and ii) an $\text{Exponential}(1)$ disribution, using the Bayesian information criterion (BIC). This is repeated 1000 $\times$ at each grid point and the two sets of BIC scores are compared using a paired t-test ($p=0.001$), which determines the colour of the point on the plot. \textit{Right:} points are instead sampled from an $\text{Exponential}(1)$ disribution. These simulations show that clock model averaging is most applicable for trees with more than 30 taxa.   } 
\label{fig:ClockModelAverging}
\end{figure}

@@

To avoid a cross-product explosion, the four targets for clock model improvement (\textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, \textbf{\nameref{sect:NER}}, and \textbf{\nameref{AVMVN_sect}}) are evaluated sequentially, in the order presented in this paper. The setting which is considered to be the best in each step is then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig. \ref{fig:tournament}}.

@@

%\textbf{Fig. \ref{fig:parameterisationResults}} shows that the \textit{real 2006} performs considerably worse than any of the other settings. This is due to the poor sampling of the prior under this setting (ie. low ESS of $p(\theta)$). The failure of \textit{real 2006} thus highlights the appeal of 1) the \textit{cat} or \textit{quant} parameterisations, both of which have trivial contributions to the prior density (ie. uniform priors), and 2) the smarter operators used by \textit{real} (Zhang and Drummond 2020). Due to its computational burden, \textit{real 2006} was not benchmarked for all of the datasets in \textbf{Table \ref{table:datasets}}.

%Our results show that the \textit{quant} parameterisation yields the best performance with respect to effective samples per hour. \textit{quant} outperforms \textit{quant 2006}, suggesting that the constant distance operators are effective. Furthermore, \textit{quant} outperforms \textit{real} especially at sampling from the posterior and prior distributions (ie. high ESS/hr for $p(\theta|D)$ and $p(\theta)$). This is most likely because of the uniform prior distribution of rate quantiles.

%Overall, \textit{quant} yields a median ESS/hr approximately 150 \% faster with respect to sampling the posterior probability, and approximately 30 \% faster with respect to sampling branch rates $r$ and clock standard deviation $\sigma$, compared with \textit{real}.

@@

\subsection*{Evaluating the AVMVN leaf rate operator}
